---
title: 'UP431 Week 5.2: Discrete Choice Analysis (2)'
author: "Jesus Barajas, Chaeyeon Han, and Bumsoo Lee"
date: "February 25, 2021"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

# Introduction

We will practice model specification, log likelihood test, and value of time calculation


# Multinomial Logit with SF MTC commute data 2000

This dataset extracted from 2000 SF MTC travel survey includes both trip- (individual-) specific variables and mode-specific variables.

caseid: case id (commute trip) 
alt: alternative number (1=Drive Alone; 2=Shared Ride 2; 3=Shared Ride 3+; 4=Transit; 5=Bike; 6=Walk)
choice: 1 if chosen this mode
numalts: number of alternatives available
ivtt: in-vehicle travel time (minutes)
ovtt: out-of-vehicle travel time
tvtt: total travel time
cost: travel cost (cents)
dist: Oneway distance to work (miles)
female: 1 if female
age
numveh: number of vehicles in household
hhinc: household income
wkccbd: 1 if working in core business district
vehbywrk: number of vehicles divided by number of workers in hh

```{r}
#install.packages("mlogit")
library(tidyverse)
library(mlogit)

SF <- read_csv("C:/Lab0/2021_UP431/Lab4/Sfmtc.csv")
head(SF, 10)

SF2 <- dfidx(SF, alt.levels= c("DA", "SR2", "SR3+", "Transit", "Bike", "Walk"))
# head(SF2, 10)
# if needed, you can also take a subset of the data, using subset argument
# SF3 <- dfidx(ModeCanada, subset= numalts==6, alt.levels= c("DA", "SR2", "SR3+", "Transit", "Bike", "Walk"))

m1.SF2 <- mlogit(choice ~ cost + tvtt | hhinc, data= SF2, reflevel= "DA")
summary(m1.SF2)

# if needed, you can do analysis only for subset of alternatives
m2.SF2 <- mlogit(choice ~ cost + tvtt | hhinc, data= SF2, alt.subset= c("DA", "SR2", "SR3+", "Transit"), reflevel= "DA")
summary(m2.SF2)

SF
```
Let's see how sensitive we are to in-vehicle vs. out-of-vehicle travel time.

```{r}
m3.SF2 <- mlogit(choice ~ cost + ivtt + ovtt | hhinc, data= SF2, alt.subset= c("DA", "SR2", "SR3+", "Transit"), reflevel= "DA")
summary(m3.SF2)
```

Let's go back to PPT

# Log Likelihood Test

McFadden R^2 of m3 is larger than that of m2. Let's perform a likelihood ratio test to see if it's statistically significantly better.

```{r}
lrtest(m3.SF2, m2.SF2)

```

the results are discussed in PPT

Export the results.

```{r}
# install.packages("stargazer")
library(stargazer)

stargazer(m2.SF2, m3.SF2, out="Output/table1.htm", type="html",
                  column.labels = c("Model 1", "Model 2"),
                  title= "Table 1: Models with cost, time, and income")
```


# Value of time
Let's go back to PPT and calculate the value of time that is manifested by commute mode choice.

## Further modification of travel time variables
Let's try to include total travel time and ovtt/dist
```{r}
SF3 <- SF %>% mutate(ovttdist= ovtt/dist)
SF4 <- dfidx(SF3, alt.levels= c("DA", "SR2", "SR3+", "Transit", "Bike", "Walk"))

m4.SF4 <- mlogit(choice ~ cost + tvtt + ovttdist | hhinc, data= SF4, alt.subset= c("DA", "SR2", "SR3+", "Transit"), reflevel= "DA")
summary(m4.SF4)

lrtest(m4.SF4, m3.SF2)
```

Modelt fit is better. Let's go back to PPT and see if value of time also makes more sense.

# Exercise
Using more relevant variables, build your best model.

```{r}
# your code here



```






