---
title: 'UP431 Lab2: Exploring NHTS Data (2)'
output:
  pdf_document: 
    latex_engine: xelatex
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<!--
Updated: 2021-02-11 by Chaeyeon Han @ University of Illinois at Urbana-Champaign UP431 Spring 2021
-->

# Using NHTS Data (2) 
  
This week, we will explore household data and vehicle data. Get ready with `tidyverse`, `haven` and NHTS SPSS dataset.

```{r}
library(tidyverse)
library(haven)
```


# Using Household Data

```{r}
hh <- read_sav("C:/Lab0/2021_UP431/Lab1/Data/spss (2)/hhpub.sav")
```

Begin with exploring what variables household data contains.

```{r}
hh <- as_factor(hh)
names(hh)
```
Filter rows that are from `Chicago-Naperville-Elgin, IL-IN-WI` CBSA. 

```{r}
chi_hh <- hh %>% filter(hh_cbsa == "Chicago-Naperville-Elgin, IL-IN-WI")
```


**Task 1** 

What do you think would be the relationship between car ownership and income group? 
What variable would you use to analyze the relationship?

Like lab 1, use `levels` and `fct_collapse` to see how income group is recorded in the dataset and customize it. 

```{r}
levels(chi_hh$hhfaminc)
```
Simplify income group to lower (< 50% of median income), moderate (50% >= and < 80%), middle (80% - 12%), and upper (>= 120%) income groups. Use 2017 median income, which is about $61,500.

```{r}
# Your code comes here
chi_hh <- chi_hh %>%
  mutate(
    hhincome_short = fct_collapse(
      hhfaminc,
      "lower" = c(
        "Less than $10,000",
        "$10,000 to $14,999",
        "$15,000 to $24,999",
        "$25,000 to $34,999"
      ),
      "moderate" = c("$35,000 to $49,999"),
      "middle" = c("$50,000 to $74,999"),
      "upper" = c(
        "$75,000 to $99,999",
        "$100,000 to $124,999",
        "$125,000 to $149,999",
        "$150,000 to $199,999",
        "$200,000 or more"
      ),
      Missing = c("I prefer not to answer", "I don't know", "Not ascertained")
    )
  ) %>% filter(hhincome_short != "Missing")
```

a) Calculate the percentage of household with no vehicle for each income group. Don't forget that every analysis should consider weight! 

```{r}
# Your code comes here
noVehicle_income <- chi_hh %>% 
  count(hhincome_short, hhvehcnt, wt = wthhfin) %>% 
  group_by(hhincome_short) %>% 
  mutate(per = prop.table(n)*100) %>%
  filter(hhvehcnt == 0) 

noVehicle_income
```
Is there any trend?


b) Calculate the mean of vehicle number for each income group. You can use `weighted.mean` to easily calculate the weighted mean.

```{r}
# Your code comes here
meanVehicle_income <- chi_hh %>% 
  group_by(hhincome_short) %>% 
  summarise(weighted_veh = weighted.mean(hhvehcnt, wthhfin))

meanVehicle_income
```

c) Calculate the mean number of vehicle to driver ratio (HHVEHCNT/WRKCOUNT), for each income group.

```{r}
# Your code comes here
meanVehicle2employee_income <- chi_hh %>% 
  group_by(hhincome_short) %>% 
  summarise(ratio = weighted.mean(hhvehcnt/drvrcnt, wthhfin)) # This arouses Inf! Guess why.

meanVehicle2employee_income
```
Why can't we calculate the ratio with the previous code?

```{r}
# Your code comes here
meanVehicle2employee_income <- chi_hh %>% 
  mutate(ratio = ifelse(drvrcnt!=0, hhvehcnt/drvrcnt, hhvehcnt)) %>% # assumed that households with no driver has 1 worker. 
  group_by(hhincome_short) %>% 
  summarise(meanRatio = weighted.mean(ratio, wthhfin))

meanVehicle2employee_income
```

d) Visualze Task 1 (a) into a bar graph using `geom_bar`. 

```{r}
# Your code comes here
noVehicle_income$per <- round(noVehicle_income$per, 2)

ggplot(noVehicle_income, aes(hhincome_short, per)) +
  geom_bar(stat = "identity", fill = "royalblue") +
  labs(x = "Income Group", y = "People with no vehicle(%)", title = "Population with No Vehicle by Income Group in Chicago CBSA") +
  geom_text(
    aes(label = paste0(per, "%"), y = per),
    vjust = 1.4,
    size = 4,
    color = "white"
  )
```
d-1) Crosstab: 
bivariate frequency tables with percent numbers. Just another way of doing a similar task.

```{r}

#install.packages("pollster")  #a package for survey analysis
library(pollster)


crosstab(df = chi_hh,
         x = hhincome_short,
         y = hhvehcnt,
         weight = wthhfin)

crosstab(
  df = chi_hh,
  x = hhincome_short,
  y = hhvehcnt,
  weight = wthhfin,
  format = "long"
) %>%
  ggplot(aes(hhincome_short, pct, fill = hhvehcnt)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Income Group", y = "People with no vehicle(%)", title = "Population with No Vehicle by Income Group in Chicago CBSA")
```


#Using Vehicle Data

Import vehicle data. Explore the variables and filter rows from Chicago-Naperville-Elgin, IL-IN-WI.

```{r}
# Your code comes here
veh <- read_sav("C:/Lab0/2021_UP431/Lab1/Data/spss (2)/vehpub.sav")
veh <- as_factor(veh)
names(veh)
chi_veh <- veh %>% filter(HH_CBSA == "Chicago-Naperville-Elgin, IL-IN-WI")
```

Vehicle data consists one row for *each* vehicle. It means that a household with three vehicles will have three rows in the vehicle data.View the dataset and check `HOUSEID` to see what it means!  

**Task 2**

Assume that you need a VMT value in a household level. You would need to aggregate the BESTMILE variable in the vehicle file to a household level using HOUSEID.Before that, make sure that there are no non-numeric values in BESTMILE.

```{r}
# Your code comes here
chi_veh <- chi_veh %>% 
  mutate(BESTMILE_new = fct_collapse(BESTMILE, Missing = c("Not ascertained"))) %>%
  filter(BESTMILE != "Missing")
```                                       

a) Use `aggregate` function. Let's leave HHFAMINC,WTHHFIN for the next task, and also HHSIZE, HBHUR, HHVEHCNT, DRVRCNT for linear regression in the last task.Name the aggregated VMT as `HHVMT`.

```{r}
# Your code comes here
agg_veh <- aggregate(as.numeric(as.character(BESTMILE_new))~HOUSEID + HHFAMINC + WTHHFIN + HBHUR + HHSIZE + HHVEHCNT + DRVRCNT, chi_veh, sum)
#agg_veh

agg_veh <- agg_veh %>% rename("HHVMT" = "as.numeric(as.character(BESTMILE_new))")
#agg_veh
```

b) Caculate the mean annual household VMT by four income group made in task 1.

```{r}
# Your code comes here
agg_veh <- agg_veh %>%
  mutate(
    HHINCOME_SHORT = fct_collapse(
      HHFAMINC,
      "lower" = c(
        "Less than $10,000",
        "$10,000 to $14,999",
        "$15,000 to $24,999",
        "$25,000 to $34,999"
      ),
      "moderate" = c("$35,000 to $49,999"),
      "middle" = c("$50,000 to $74,999"),
      "upper" = c(
        "$75,000 to $99,999",
        "$100,000 to $124,999",
        "$125,000 to $149,999",
        "$150,000 to $199,999",
        "$200,000 or more"
      ),
      Missing = c("I prefer not to answer", "I don't know", "Not ascertained")
    )
  ) %>% 
  filter(HHINCOME_SHORT != "Missing")
```

```{r}
# Your code comes here
meanHHVMT_income <- agg_veh %>% 
  group_by(HHINCOME_SHORT) %>% 
  summarise(HHVMT = weighted.mean(HHVMT, WTHHFIN))

meanHHVMT_income
```

c) Make an ANOVA test to check the relationship between annual household VMT and income group.

```{r}
# Your code comes here
library(car)

leveneTest(agg_veh$HHVMT, agg_veh$HHINCOME_SHORT)
oneway.test(agg_veh$HHVMT~agg_veh$HHINCOME_SHORT, var.equal=F)
```
**Task 3** 

Lastly, conduct a linear regression analysis at the household level.

HHVMT = f(HHSIZE, HHVEHCNT/WRKCOUNT, INCGROUP --> dummy, HBHUR --> dummy)

a) Create dummy variables (https://www.marsja.se/create-dummy-variables-in-r/).
You can either use a libary to make it fast, or you can make dummy variables by yourself using `ifelse`.

```{r}
#install.packages("fastDummies")
library(fastDummies)
```

If you want to use `fastDummies`, rename any variable with an empty space(' ') in the name, since you will not be able to access the variable by name if there is a space.
 
```{r}
# Your code comes here
agg_veh <- agg_veh %>%
  mutate(HBHUR_new = fct_collapse(
    HBHUR,
    "SmallTown" = c("Small Town"),
    "SecondCity" = c("Second City"),
    "Missing" = c("Not ascertained")
  )) %>%
  filter(HBHUR_new != "Missing")
```

```{r}
# Your code comes here
agg_veh <- dummy_cols(agg_veh, select_columns = c("HHINCOME_SHORT","HBHUR_new"))
#agg_veh
```

a-2) IF you want to use `ifelse`, here is an example.

```{r}
#exampleData$Var_A <- ifelse(exampleData$Var == 'A', 1, 0)
#exampleData$Var_B <- ifelse(exampleData$Var == 'B', 1, 0)
```

b) Run a regression and print the result using `summary` function.

```{r}
# Your code comes here
HH_model <- lm(HHVMT ~ HHSIZE + HHVEHCNT/DRVRCNT + HHINCOME_SHORT_lower + HHINCOME_SHORT_moderate + HHINCOME_SHORT_upper + HBHUR_new_Urban + HBHUR_new_SmallTown + HBHUR_new_Suburban + HBHUR_new_Rural, data = agg_veh, weights = WTHHFIN)

summary(HH_model)
```
Do you see any significant relationship?

```{r}
aov(HH_model)
```

c) What additional variables do you want to include in the regression?

```{r}
# Your code comes here
```


&nbsp;
