---
title: 'UP431 Lab2: Exploring NHTS Data (2)'
output:
  html_document:
    df_print: paged
  pdf_document: 
    latex_engine: xelatex
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<!--
Updated: 2021-02-11 by Chaeyeon Han @ University of Illinois at Urbana-Champaign UP431 Spring 2021
-->

# Using NHTS Data (2) 
  
This week, we will explore household data and vehicle data. Get ready with `tidyverse`, `haven` and NHTS SPSS dataset.

```{r}
library(tidyverse)
library(haven)
```


# Using Household Data

```{r}
hh <- read_sav("C:/Lab0/2021_UP431/Lab1/Data/spss (2)/hhpub.sav")
```

Begin with exploring what variables household data contains.

```{r}
hh <- as_factor(hh)
names(hh)
```
Filter rows that are from `Chicago-Naperville-Elgin, IL-IN-WI` CBSA. 

```{r}
chi_hh <- hh %>% filter(hh_cbsa == "Chicago-Naperville-Elgin, IL-IN-WI")
```


**Task 1** 

What do you think would be the relationship between car ownership and income group? 
What variable would you use to analyze the relationship?

Like lab 1, use `levels` and `fct_collapse` to see how income group is recorded in the dataset and customize it. 

```{r}
levels(chi_hh$hhfaminc)
```

```{r}
# Your code comes here

```

a) Calculate the percentage of household with no vehicle for each income group. Don't forget that every analysis should consider weight! 

```{r}
# Your code comes here

```
Is there any trend?


b) Calculate the mean of vehicle number for each income group. You can use `weighted.mean` to easily calculate the weighted mean.

```{r}
# Your code comes here

```

c) Calculate the mean number of vehicle to driver ratio (HHVEHCNT/WRKCOUNT), for each income group.

```{r}
# Your code comes here

```
Why can't we calculate the ratio with the previous code?

```{r}
# Your code comes here

```

d) Visualze Task 1 (a) into a bar graph using `geom_bar`. 

```{r}
# Your code comes here

```

d-1) Crosstab: 
bivariate frequency tables with percent numbers. Just another way of doing a similar task.

```{r, eval=FALSE}
#install.packages("pollster")  #a package for survey analysis
library(pollster)


crosstab(df = chi_hh,
         x = hhincome_short,
         y = hhvehcnt,
         weight = wthhfin)

crosstab(
  df = chi_hh,
  x = hhincome_short,
  y = hhvehcnt,
  weight = wthhfin,
  format = "long"
) %>%
  ggplot(aes(hhincome_short, pct, fill = hhvehcnt)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Income Group", y = "People with no vehicle(%)", title = "Population with No Vehicle by Income Group in Chicago CBSA")
```


#Using Vehicle Data

Import vehicle data. Explore the variables and filter rows from Chicago-Naperville-Elgin, IL-IN-WI.

```{r, eval = FALSE}
veh <- read_sav("C:/Lab0/2021_UP431/Lab1/Data/spss (2)/vehpub.sav") # your path
veh <- as_factor(veh)
names(veh)
chi_veh <- veh %>% filter(HH_CBSA == "Chicago-Naperville-Elgin, IL-IN-WI")
```

Vehicle data consists one row for *each* vehicle. It means that a household with three vehicles will have three rows in the vehicle data.View the dataset and check `HOUSEID` to see what it means!  

**Task 2**

Assume that you need a VMT value in a household level. You would need to aggregate the BESTMILE variable in the vehicle file to a household level using HOUSEID.Before that, make sure that there are no non-numeric values in BESTMILE.

```{r, eval = FALSE}
chi_veh <- chi_veh %>% 
  mutate(BESTMILE_new = fct_collapse(BESTMILE, Missing = c("Not ascertained"))) %>%
  filter(BESTMILE != "Missing")
```                                       

a) Use `aggregate` function. Let's leave HHFAMINC,WTHHFIN for the next task, and also HHSIZE, HBHUR, HHVEHCNT, DRVRCNT for linear regression in the last task.Name the aggregated VMT as `HHVMT`.

```{r}
# Your code comes here

```

b) Caculate the mean annual household VMT by four income group made in task 1.

```{r}
# Your code comes here

```

```{r}
# Your code comes here

```

c) Make an ANOVA test to check the relationship between annual household VMT and income group.

```{r}
library(car)

# Your code comes here
```
**Task 3** 

Lastly, conduct a linear regression analysis at the household level.

HHVMT = f(HHSIZE, HHVEHCNT/WRKCOUNT, INCGROUP --> dummy, HBHUR --> dummy)

a) Create dummy variables (https://www.marsja.se/create-dummy-variables-in-r/).
You can either use a libary to make it fast, or you can make dummy variables by yourself using `ifelse`.

```{r}
#install.packages("fastDummies")
library(fastDummies)
```

If you want to use `fastDummies`, rename any variable with an empty space(' ') in the name, since you will not be able to access the variable by name if there is a space.
 
```{r, eval = FALSE}
agg_veh <- agg_veh %>%
  mutate(HBHUR_new = fct_collapse(
    HBHUR,
    "SmallTown" = c("Small Town"),
    "SecondCity" = c("Second City"),
    "Missing" = c("Not ascertained")
  )) %>%
  filter(HBHUR_new != "Missing")
```

```{r, eval = FALSE}
# Your code comes here
agg_veh <- dummy_cols(agg_veh, select_columns = c("HHINCOME_SHORT","HBHUR_new"))
#agg_veh
```

a-2) IF you want to use `ifelse`, here is an example.

```{r}
#exampleData$Var_A <- ifelse(exampleData$Var == 'A', 1, 0)
#exampleData$Var_B <- ifelse(exampleData$Var == 'B', 1, 0)
```

b) Run a regression and print the result using `summary` function.

```{r}
# Your code comes here

```
Do you see any significant relationship?

c) What additional variables do you want to include in the regression?

```{r}
# Your code comes here
```


&nbsp;
